Abstractions for Multi-Sorted Substitutions

### 2023-08-22 paper/paper.tex
Update.

  \subsection{Multi-Sorted Syntax}
  To avoid these drawbacks, we define the whole syntax as a single sort-indexed type of terms.

  A sort describes to which syntactic category a term belongs and is
  itself indexed by a sort type which describes whether the syntax
  permits variables of this sort:
  \LibCode\KSortTy
  \AppCode\FSort

  The term type \texttt{S~⊢~s} is then indexed by its sort \texttt{s} and the sorts of its free variables \texttt{S}.
  For example, \texttt{[𝕥, 𝕥] ⊢ 𝕖} is the type of expressions (\texttt{𝕖}) with two free type-variables (\texttt{𝕥}).
  \newpage
  \AppCode\FSyntax
  The notation \texttt{\_⊢\_} is often used for well-typed terms in
  intrinsically-typed languages. This is no accident: in effect, we
  defined an intrinsically-typed language with the twist that the
  typing relation assures exactly that the syntactic categories are followed.
  Sorts \texttt{s} correspond to types, and lists of sorts \texttt{S}
  correspond to type environments.

@@

  \subsection{A Structure for Multi-Sorted Syntax}
  The rest of our framework builds on top of a multi-sorted
  syntax for which we define the following structure:
  \LibCode\KTerms
  The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \texttt{`\_} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FTerms

  \section{Renamings \& Substitutions}
  \subsection{Multi-Sorted Renamings \& Substitutions}
  Working with a sort-indexed syntax allows us to define renamings and substitutions
  which replace all variables of all sorts simultaneously:

@@

  \subsection{Unifying Renamings \& Substitutions}
  A well-known problem with DeBruijn-indices is that the operation of
  applying a substitution \texttt{σ} to a term \texttt{t} cannot be
  defined directly by structural recursion on \texttt{t}.
  If \texttt{t} is a term like \texttt{λx e},
  where the subterm \texttt{e} is under a binder, then the variables
  in \texttt{e} are shifted, i.e.  the variable \texttt{zero} refers
  to the variable from the binder, whereas a variable \texttt{suc x}
  refers to the variable \texttt{x} from outside of the binder.
  As the terms contained in \texttt{σ} may themselves contain free variables,
  they need to be shifted, too, before they can be substituted
  for variables in \texttt{e}. 
  Shifting the variables in a term is itself a substitution, which maps
  each variable \texttt{x} to the term \texttt{` (suc x)}.
  However, applying this shifting substitution to the terms in
  \texttt{σ} is not structurally recursive, as the terms in \texttt{σ}
  are not subterms of \texttt{λx e}.

  The usual workaround is to define renamings first and use them
  to weaken the terms in substitutions when going under binders.
  While this keeps lemmas simple (requiring only structural induction),
  it duplicates the amount of work as all operations and lemmas have
  to be defined and proved for both renamings and substitutions.

  To avoid this duplication,
  McBride\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
  introduced an abstraction called \emph{Kits}, which allows to talk about
  both renamings and substitutions simultaneously.

  The core observation leading to this abstraction is that renamings
  and substitutions only differ in their return type:
  \ExampleCode\FExampleSubRen

@@

@@

>>>>

  \subsection{Multi-Sorted Syntax}
  To avoid these drawbacks, we define the syntax as a single
  sort-indexed type of terms.

  A sort describes to which syntactic category a term belongs and is
  itself indexed by a sort type, which describes whether the syntax
  permits variables of this sort:
  \LibCode\KSortTy
  \AppCode\FSort

  The term type \texttt{S~⊢~s} is then indexed by its sort \texttt{s} and the sorts of its free variables \texttt{S}.
  For example, \texttt{[𝕥, 𝕥] ⊢ 𝕖} is the type of expressions (\texttt{𝕖}) with two free type-variables (\texttt{𝕥}).
  \newpage
  \AppCode\FSyntax
  The notation \texttt{\_⊢\_} is often used for terms in
  intrinsically-typed languages. This is no accident: in effect, we
  defined an intrinsically-typed language with the twist that the
  typing relation assures exactly that the syntactic categories are followed.
  Sorts \texttt{s} correspond to types, and lists of sorts \texttt{S}
  correspond to type environments.

@@

  \subsection{A Structure for Multi-Sorted Syntax}
  The multi-sorted syntax makes it easy to capture general syntax definitions
  in a structure:
  \LibCode\KTerms
  The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \texttt{`\_} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FTerms

  \section{Renamings \& Substitutions}
  \subsection{Multi-Sorted Renamings \& Substitutions}
  Working with a sort-indexed syntax allows us to define renamings and substitutions
  that replace all variables of all sorts simultaneously:

@@

  \subsection{The Need for Renamings}
  A problem with mechanizing languages with DeBruijn-indices is that
  the operation of applying a substitution \texttt{σ} to a term
  \texttt{t} cannot be defined directly by structural recursion on
  \texttt{t}.
  If \texttt{t} is a term like \texttt{λx e},
  where the subterm \texttt{e} is under a binder, then the variables
  in \texttt{e} are shifted, i.e.\ the variable \texttt{zero} refers
  to the variable from the binder, whereas a variable \texttt{suc x}
  refers to the variable \texttt{x} from outside of the binder.
  As the terms contained in \texttt{σ} may themselves contain free variables,
  they need to be shifted, too, before they can be substituted
  for variables in \texttt{e}. 
  Shifting the variables in a term is itself a substitution, which maps
  each variable \texttt{x} to the term \texttt{` (suc x)}.
  However, applying this shifting substitution to the terms in
  \texttt{σ} is not structurally recursive, as the terms in \texttt{σ}
  are not subterms of \texttt{λx e}.

  The usual workaround is to first define renamings,
  i.e.\ substitutions that replace variables with variables.
  In contrast to general substitutions, renamings can be easily
  shifted: they only contain variables \texttt{x}, which can be
  shifted by simply taking their successor \texttt{suc x}.
  As the shifting substitution is a renaming, we can then define substitution,
  where we apply the shifting \emph{renaming} to the terms in \texttt{σ} when
  the substitution needs to go under a binder.

  % The usual workaround is to define renamings first and use them
  % to weaken the terms in substitutions when going under binders.
  While this keeps lemmas simple (requiring only structural induction),
  it duplicates the amount of work as all operations and lemmas have
  to be defined and proved for both renamings and substitutions.

  \subsection{Unifying Renamings \& Substitutions}
  To avoid the duplication between renamings and substitutions,
  McBride\cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}
  introduced an abstraction called \emph{Kits}.

  % While Kits were originally formulated for intrinsically-typed languages,
  % they nicely extend to our extrinsically-typed multi-sorted syntax, whose
  % sort-indexing is a form of intrinsic typing.

  A kit is a structure that abstracts over the difference between variables and terms.
  There are always exactly two instances for Kits, one for variables and one for terms,

  The core observation leading to this abstraction is that renamings
  \texttt{S₁ →ᵣ S₂}
  and substitutions \texttt{S₁ →ₛ S₂} only differ in their return type:
  while a substitution returns a term \texttt{S₂ ⊢ s}, a renaming returns
  a variable \texttt{S₂ ∋ s}.

  A Kit is a structure that abstracts precisely over the commonalities
  between variables and terms.

  % \ExampleCode\FExampleSubRen

@@

  \section{Reflection \& Generics}

  \section{Future Work}

@@

  \section{Conclusion}
